Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
ref cheng GUI paper
update data structures section

The graphics literature [@Swayne1998; @Unwin1996; @VIM; @MissingDataGUI; ] provides several solutions for exploring missings visually. These approaches incorporate missings into the plot in some way. For example, imputing values 10% below the range to display all values in a scatter plot. These methods require the data to be linked to the missing or imputed values; this link is typically hidden from the user. The approaches from the graphics literature have not been translated and integrated into tidy data. This limits the analysts capacity to efficiently explore missing values. Extending tidy data and tidy tools to account for missing values would create more efficient and robust workflows and analysis.

@@

To explore missing data there needs to be a good representation of missing
values that integrate into analysis pipelines. A common representation of
missing values is a binary matrix the same dimension as the data, where 0 and 1
indicate not missing and missing. This is typically referred to as a matrix $R$,
used to describe missingness dependence patterns, (see for example @Rubin1976
and @Schafer2002). This matrix is used to explore missing values in the
interactive graphics library XGobi, where it was called a "shadow matrix",
defined as a copy of the original data with indicator values of missingness:
"NA" and "!NA" for missing and not missing, represented as 0 and 1 (see Figure
\@ref(fig:shadow-matrix-progress) parts 2 and 3 below).

>>>>

The graphics literature [@Swayne1998; @Unwin1996; @VIM; @Cheng2015] provides several solutions for exploring missings visually. These approaches incorporate missings into the plot in some way. For example, imputing values 10% below the range to display all values in a scatter plot. These methods require the data to be linked to the missing or imputed values; this link is typically hidden from the user. The approaches from the graphics literature have not been translated and integrated into tidy data. This limits the analysts capacity to efficiently explore missing values. Extending tidy data and tidy tools to account for missing values would create more efficient and robust workflows and analysis.

@@

To explore missing data there needs to be a good representation of missing
values that integrate into analysis pipelines. The R matrix, where 0 and 1
indicate not missing and missing provides a template, as it is very common in
missing data literature.  This matrix has been used to explore missing values in
the interactive graphics library XGobi, where it was called a "missing value
shadow", or "shadow matrix", defined as a copy of the original data with
indicator values of missingness, where 1s represent missing values, and 0
represents complete values. The values 0 and 1 (or equally, FALSE and TRUE) can
be confusing, however, as it is not immediately clear what they represent - does
0 indicate a the absence of observation, or the presence of a missing value? The
shadow matrix was specially linked to the data, but could not be referred to
directly, instead they were linked with special window views. This meant that
the data structure was abstracted from the user to facilitate data exploration.
This made sense at the time with the state of the art for graphicaly analysis,
but that data structure needs to be clearly presented to the user so that they
can more broadly use it in other data analyses.
